---
title: "Data Exploration"
author: "Bruno Caneco"
toc: true
toc-depth: 4
format: 
  html:
    df-print: paged
    code-fold: true
editor: visual
---

```{r}
#| include: false

library(tidyverse)
library(readxl)
library(MetBrewer)

theme_set(theme_bw())

col_pal <- met.brewer("Lakota", n = 2)
```




Rough and overall look at the data to get familiar with variables, basic
trends and spot any odd values.


## Data description

- The dataset contains acoustic data of calls from two subjects, *Riet* and *Alex*. 
- Data was recorded over two separate sessions for each subject.
- Rows correspond to call events and columns provide acoustic features of each call

```{r}
# load data
data <- read_xlsx(path = "data/New Master Excel.xlsx")
data
```

### Variables of interest

Variables can be broadly grouped into:

- Call features:
  - `Delta Time`: duration of the call (secs)
  - `Max Freq`: maximum frequency of the call (Hz)
  - `PFC Avg Slope`: voice inflection of the call
  - `Avg Entropy`: entropy of the call (measuring efficiency in transmitting a message?)
  
- Change from preceding call
  - `ΔVoice Freq`: in maximum frequency
  - `ΔVoice Inflexion`: in voice inflection
  -  `ΔVoice Entropy`: in entropy



## Data pre-processing

- Adding a couple of new columns:
  - `delta_duration`: change in duration of two consecutive calls (as per chat with Adriano)
  - `session`: within-subject session identifier

- Modify column names for easier coding


```{r}
data <- data %>% 
  group_by(File) %>%
  # add column with change in call duration
  mutate(
    delta_duration = `Delta Time` - lag(`Delta Time`), 
    .after = `Delta Time`
  ) %>%
  group_by(Subject) %>%
  # add session number
  mutate(
    session = factor(File, levels = unique(File), labels = 1:length(unique(File))),
    .after = Subject
    ) %>%
  ungroup() %>%
  # rename columns
  rename_with(.fn = ~ str_replace(., pattern = "Δ", replacement = "delta_")) %>%
  rename_with(.fn = ~ str_to_lower(str_replace_all(., pattern = " ", replacement = "_")))

#head(data)

```


```{r}
#| include: false

# save data out
write_csv(x = data, file = "data/calls_data.csv")
```


## Exploring data at subject and session level


### Overall (naive) data summaries

```{r}
data %>%
  group_by(subject, session) %>%
  summarise(
    n_calls = n(),   # nr calls/session
    session_duration = max(end_time), # session duration
    #calls_per_min = n_calls/session_duration * 60, # nr calls/minute
    across(
      c(delta_time, low_freq, high_freq, max_freq, pfc_avg_slope, 
        avg_entropy, agg_entropy, snr, delta_voice_freq, delta_voice_inflexion,
        delta_voice_entropy), 
      mean, 
      .names = "mean_{.col}",
      na.rm = TRUE,
    ),
    .groups = "drop"
  )
```

Obvious takeaways from per subject and session **naive** summaries:

- *Alex* produced similar number of calls in the two sessions, which have
similar recording duration.
- *Riet*'s sessions quite dissimilar in terms of number of calls relative to
recording duration. Further exploration will reveal the reason(s) behind this.
- Calls produced by *Riet* last longer than those of *Alex*.
- *Riet* calls show, on average, lower levels of low frequency and maximum frequency
than *Alex*.
- No substantial differences (at least for a non-specialist!) between sessions
and subjects in the remaining acoustic variables.


### Call-specific measurements through recording time

```{r}
#| fig-width: 8
#| fig-height: 12

data %>%
  select(subject:begin_time, low_freq, high_freq, max_freq:snr) %>%
  pivot_longer(cols = -c(subject, session, begin_time), names_to = "variable") %>%
  ggplot() +
  geom_line(aes(x = begin_time, y = value, color = session)) +
#  geom_point(aes(x = begin_time, y = value, color = session)) +
  facet_grid(variable~subject, scales = "free") +
  scale_color_manual(values = col_pal)

```

- Discrepancies found earlier between *Riet*'s sessions in nr. of calls
relative to recording duration explained by a large time gap between the first
few calls and subsequent calls during session 1.
- Otherwise, no clear differences between sessions for the displayed variables.
- *Alex*'s calls appear to show larger variability in `high_freq`,
`low_freq`, `pfc_avg_slope` and `sur`.




### Call acoustic properties vs. call duration

```{r}
#| fig-width: 8
#| fig-height: 12

data %>%
  select(subject, session, delta_time, low_freq, high_freq, max_freq:snr) %>%
  pivot_longer(cols = -c(subject, session, delta_time), names_to = "variable") %>%
  ggplot() +
  geom_point(aes(x = delta_time, y = value, color = session)) +
  facet_grid(variable~subject, scales = "free_y") +
  scale_color_manual(values = col_pal)

```

- No obvious relationship between displayed variables and call duration
- *Alex*'s shorter call duration is highlighted again, as well as the wider
distribution of `high_freq`, `low_freq`, `pfc_avg_slope` and `sur` values.



### Change in call duration vs. change in entropy, frequency and inflection

```{r}
#| fig-width: 8
#| fig-height: 6
data %>%
  select(subject, session, delta_duration, delta_voice_freq:delta_voice_entropy) %>%
  drop_na() %>%
  pivot_longer(cols = -c(subject, session, delta_duration), names_to = "variable") %>%
  ggplot() +
  geom_point(aes(x = delta_duration, y = value, color = session)) +
  facet_grid(variable~subject, scales = "free_y") +
  scale_color_manual(values = col_pal)
```

- No obvious trend between change in call duration and change in entropy,
frequency or inflection.
- Change in call duration between consecutive calls less variable in
*Alex*'s calls than those from *Riet* (as expected given *Alex*'s shorter calls).
- Change in voice inflection between consecutive calls is less variable in
*Riet*'s calls.


### Slope Vs Entropy

```{r}
#| fig-width: 10
#| fig-height: 4

data %>%
  ggplot(aes(x = avg_entropy, y = pfc_avg_slope, color = session)) +
  geom_point() +
  facet_wrap(~subject) +
  scale_color_manual(values = col_pal)
```

- No clear relationship between slope (voice inflection) and entropy.
